Trauma treatment delay

### 2023-09-18 manuscriptlongversion.Rmd
intro

Within the realm of modern medicine, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients. The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020]. As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological. The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body. It is often divided into two subgroups, blunt force trauma and penetrative trauma. Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of this could be car crashes, falls or direct blows to the body. Penetrative trauma is when an object pierces the skin or body and creates an open wound. Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020]. Differences exist between the two groups, as blunt trauma patients seem to score higher on ISS scale on arrival, require more resources and be hospitalized for a longer period [@FITCH20191201].

@@

Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012].

However, this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. According to the WHO, about 90% of all trauma related deaths occur in low- and middle-income countries, with injury deaths rates being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries. Problems identified with trauma care within these countries were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

>>>>

Within the realm of modern medicine, trauma care stands as an important part in healthcare and is crucial to lowering mortality and morbidity in injured patients. The definition of medical trauma is "a set of psychological and physiological responses to pain, injury, serious illness, medical procedures and frightening treatment experiences" [@International_Society_for_Traumatic_Stress_Studies_2020]. As the definition suggests, it encompasses a broad spectrum of different conditions, both physical and psychological. The one we will be focusing on in this paper is physical trauma, which is defined as a serious injury to the body. It is often divided into two subgroups, blunt force trauma and penetrative trauma. Blunt force trauma happens when an object or force strikes the body, often causing bruising, broken bones or deep cuts. Examples of this could be car crashes, falls or direct blows to the body. Penetrative trauma is when an object pierces the skin or body and creates an open wound. Examples of this are gunshot wounds and stab wounds [@National_Institute_of_General_Medical_Sciences_2020]. Differences exist between the two groups, as blunt trauma patients tend to score higher on ISS scale on arrival and as such require more resources and be hospitalized for a longer period because of their more sever injuries[@FITCH20191201].

@@

Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. Studies estimate the cost of trauma care being between 18,500\$ and 41,500\$ in HIC, depending on the country [@Willenberg_Curtis_Taylor_Jan_Glass_Myburgh_2012]. However, as one might expect, this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. It is not only patient level factors that affect how people might be affected by trauma. About 90% of all trauma related deaths occur in low- and middle-income countries, with death rates by trauma also being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries. Problems identified within these countries that contribute to these statisitics were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].
